---
title: "Week1_Ch2"
author: "Sahana Ramakrishnan"
date: "2024-07-18"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Chapter 2
### 2.3.1 Basic Commands

```{r}
x <- c(1,3,2,5)
x
```

```{r}
x = c(1,6,2)
x
y = c(1,4,3)
y
```

#### Using functions

```{r}
length(x)
length(y)
x+y
```

#### Using ls() function

```{r}
ls()
rm(x,y)
ls()

```
#### removing all objects at once
```{r}
 rm(list=ls())
```


#### Creating matrix

```{r}
?matrix
x = matrix(data=c(1,2,3,4), nrow=2, ncol=2)
x
```


```{r}
x = matrix(c(1,2,3,4) ,2,2)
matrix (c(1,2,3,4) ,2,2,byrow=TRUE)
```

#### Using sqrt() function
```{r}
sqrt(x)
x^2
```

#### rnorm() for creating vectors
```{r}
x = rnorm(50)
y = x + rnorm(50,mean=50,sd=.1)
cor(x,y)

```

```{r}
set.seed(1303)
rnorm(50)
```

```{r}
set.seed(3)
y = rnorm(100)
mean(y)
var(y)
sqrt(var(y))
sd(y)
```
### 2.3.2 Graphics

#### Using plot functions

```{r}
x = rnorm(100)
y = rnorm(100)
plot(x,y)
plot(x, y, xlab = "this is the x-axis", ylab = "this is the y-axis",
main = "Plot of X vs Y")


```

#### Using pdf() filetype
```{r}
pdf("Figure.pdf")
plot(x,y,col="green")
dev.off()       # denoting that we are done creating plot
```
#### Creating sequence of numbers
```{r}
x = seq(1, 10)
x
x = 1:10
x = seq(-pi,pi,length =50)
x
```

#### Using contour function - to finetune the output

```{r}
y = x
f = outer(x,y,function (x,y)cos(y)/(1+x^2))
contour(x,y,f)
contour (x,y,f,nlevels =45,add=T)
fa = (f-t(f))/2
contour (x,y,fa,nlevels =15)
```
#### Using image function like contour but with color plots

```{r}
image(x,y,fa)
persp(x,y,fa)
persp(x,y,fa,theta =30)
persp(x,y,fa,theta =30,phi =20)
persp(x,y,fa,theta =30,phi =70)
persp(x,y,fa,theta =30,phi =40)
```

- persp: to produce 3-dimensional plot

### 2.3.3 Indexing Data

```{r}
A = matrix (1:16,4,4)
A
```

```{r}
A[2,3]
```


```{r}
A[c(1,3),c(2,4)]
A[1:3,2:4]
A[1:2,]
A[,1:2]
```
#### Retrieving with no rows/no columns

```{r}
A[1,]
A[-c(1,3) ,]  # to keep all rows/columns except the ones indicated in index

```
#### Using dim function

```{r}
dim(A)
```

### 2.3.4 Loading Data

```{r}
# To load Auto.data file into R
Auto = read.table("Auto.data")
fix(Auto)  # to view in spreadsheet format
```

#### To clean the data for missing elements

```{r}
Auto = read.table("Auto.data",header=T,na.strings ="?")
fix(Auto)
```

#### Using read.csv for auto.csv file

```{r}
Auto = read.csv("Auto.csv",header=T,na.strings ="?")
fix(Auto)
dim(Auto)
Auto[1:4,]
```

#### Using na.omit() to remove missing data rows

```{r}
Auto = na.omit(Auto)
dim(Auto)
```

#### To check variable names

```{r}
names(Auto)
```

### 2.3.5 Additional Graphical & Numerical summaries

```{r}
# scatterplot of quantifying variables
## plot(cylinders , mpg) 
```
- Throws an error: since the variable name in data set should be accessed using "$"

```{r}
plot(Auto$cylinders , Auto$mpg )
attach(Auto)
plot(cylinders , mpg)
```

#### Using cylinder as qualitative variable since there are less no. of values

```{r}
cylinders = as.factor(cylinders )
cylinders
```

#### plotting the variable

```{r}
plot(cylinders , mpg)
plot(cylinders , mpg , col ="red ")
plot(cylinders , mpg , col ="red", varwidth =T)
plot(cylinders , mpg , col ="red", varwidth =T,horizontal =T)
plot(cylinders , mpg , col ="red", varwidth =T, xlab="cylinders ", ylab="MPG")
```

- Since the x-axis values are categorical, the plots will are boxplots.

#### Plotting histograms

```{r}
hist(mpg)
hist(mpg ,col =2)
hist(mpg ,col=2, breaks =15)
```
#### Using pairs() function to create scatterplot matrix


```{r}
# since pairs(auto) throws an error due to horsepower column being non-numeric
Auto <- read.csv("Auto.csv", stringsAsFactors = FALSE)

# Converting "horsepower" to numeric, forcing non-numeric values to be NA
Auto$horsepower <- as.numeric(Auto$horsepower)

# Filling NA values in 'horsepower' with the mean of the column
mean_hp <- mean(Auto$horsepower, na.rm = TRUE)
Auto$horsepower[is.na(Auto$horsepower)] <- mean_hp

# selecting only numeric columns
numeric_columns <- sapply(Auto, is.numeric)
Auto_numeric <- Auto[, numeric_columns]


# now creating pairs
pairs(Auto_numeric)
pairs(~ mpg + displacement + horsepower + weight + acceleration , Auto_numeric)
```
#### Identifying value of a particular variable from plot

```{r}
plot(horsepower ,mpg)
identify (horsepower ,mpg ,name)
```
#### Creating summary of Auto

```{r}
summary(Auto)
summary(mpg)

```













