---
title: "Week2_Ch2_Exercise"
author: "Sahana Ramakrishnan"
date: "2024-07-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```
## Chapter 2
### Exercise 2.4
### 8. a)
```{r}
#Setting working directory
setwd("D:/NewEng/Summer 2024/Week 2/R/")

#Reading the csv file
college <- read.csv("College.csv")

```

### 8. b)

```{r}
#Viewing the data
#View(college)

#since first column is just names of the university
rownames(college) <- college[, 1]
View(college)

```

```{r}
# to eliminate first column in data where names are stored
college <- college [, -1]
View(college)
```

### 8. c) i)
```{r}
#Summary function to see numerical summary of the data
summary(college)

```
### 8. c) ii)
```{r}
# to produce scatterplot matrix of first 10 columns
college$Private <- college$Private == "Yes"
pairs(college[, 1:10])
```

### 8. c) iii)

```{r}
# to produce side-by-side boxplots of Outstate vs Private
plot(college$Private, college$Outstate,
     xlab = "Private",
     ylab = "OutState",
     main = "Boxplot of OutState Vs Private")
```
### 8. c) iv)

```{r}
# Creating new qualitative variable: elite

Elite <- rep("No", nrow(college))
Elite[college$Top10perc > 50] <- "Yes"
Elite <- as.factor(Elite)
college <- data.frame(college, Elite)

```

```{r}
#Summary function to see how many elite universities there are
summary(college$Elite)
```

```{r}
#plot() function to produce boxplot: Outstate vs Elite

plot(college$Elite, college$Outstate,
     xlab = "Elite",
     ylab = "Outstate",
     main = "Boxplot of Outstate vs Elite")
```
### 8. c) v)

```{r}
# to divide plotting window into 2x2 grid
par(mfrow = c(2, 2))
```

```{r}
# Creating histograms for a few quantitative variables with different numbers of bins
hist(college$Apps, breaks = 10, main = "Applications Histogram", xlab = "No. of Applications", col = "blue")
hist(college$Enroll, breaks = 15, main = "Enrollments Histogram", xlab = "No. of Enrollments", col = "green")
hist(college$Outstate, breaks = 20, main = "OutState Histogram", xlab = "OutState", col = "red")
hist(college$PhD, breaks = 12, main = "PhD Faculty Histogram", xlab = "PhD Faculty %", col = "purple")
```
### 8. c) vi) Summary of the stats

```{r}
summary(college)

# correlation matrix
library(corrplot)
correlation_matrix <- cor(college[, sapply(college, is.numeric)])
corrplot(correlation_matrix, method = "circle")
```

```{r}
# box plot for key variables: Graduation Rate by Private/Public

plot(college$Private, college$Grad.Rate, 
     xlab = "Private", 
     ylab = "Graduation Rate", 
     main = "Graduation Rate by Private/Public")

# for Alumni donation % vs Elite
plot(college$Elite, college$perc.alumni, 
     xlab = "Elite University", 
     ylab = "Alumni Donation Percentage", 
     main = "Alumni donation % vs Elite")
```
- Private institutions generally have higher Outstate tuition compared to public institutions, as observed in the boxplots.
- Private institutions also tend to have higher graduation rates compared to public institutions.
- Elite universities show a higher % of alumni donations compared to non-elite universities.
- There is a strong positive correlation between the no. of full-time undergraduates and the no. of applications received.

These observations provide insights into the characteristics and differences between different types of institutions, their costs, and the demographic and academic profiles of their students

### 9. a)

#### Auto dataset

```{r}
# reading the data
auto <- read.csv("Auto.csv", na.strings = "?")

```

```{r}
# Removing rows with missing values
auto <- na.omit(auto)
#auto
```
- Quantitative predictors: mpg, cylinders, displacement, horsepower, weight, acceleration, year which are numeric and can be used for arithmetic calculations
- Qualitative predictors: origin, name which are categorical and represents groups or categories

### 9. b)

```{r}
# to determine range of each quantitative predictor 

# Calculating range for each quant. predictors
quantitative_vars <- c("mpg", "cylinders", "displacement", "horsepower", "weight", "acceleration", "year")

# Looping through each quant. variable
for (var in quantitative_vars) {
  cat(var, ":", range(auto[[var]]), "\n")
}
```
- range(auto[[var]]) calculates the range (min and max values) of the specified variable.

### 9. c) 

```{r}
# Looping through the above quant. variable and finding mean and SD
for (var in quantitative_vars) {
  cat(var, "Mean:", mean(auto[[var]]), "SD:", sd(auto[[var]]), "\n")
}
```
### 9. d)

```{r}
# removing 10th through 85th observation
auto_85 <- auto[-(10:85), ]

head(auto_85)
dim(auto_85)
```

```{r}
# calculating mean, range and SD of the remaining subset of data 

for (var in quantitative_vars) {
  cat(var, "\n")
  cat("  Range:", range(auto_85[[var]]), "\n")
  cat("  Mean:", mean(auto_85[[var]]), "\n")
  cat("  SD:", sd(auto_85[[var]]), "\n")
}
```

### 9. e)

```{r}
# Setting 2x2 plotting window
par(mfrow = c(2, 2))

# mpg vs. horsepower
plot(auto$horsepower, auto$mpg,
     xlab = "Horsepower",
     ylab = "MPG",
     main = "Horsepower vs. MPG",
     col = "blue")

# weight vs. mpg
plot(auto$weight, auto$mpg,
     xlab = "Weight",
     ylab = "MPG",
     main = "Weight vs. MPG",
     col = "green")

# displacement vs. mpg
plot(auto$displacement, auto$mpg,
     xlab = "Displacement",
     ylab = "MPG",
     main = "Displacement vs. MPG",
     col = "red")

# acceleration vs. mpg
plot(auto$acceleration, auto$mpg,
     xlab = "Acceleration",
     ylab = "MPG",
     main = "Acceleration vs. MPG",
     col = "purple")

# Resetting plotting window
par(mfrow = c(1, 1))
```

```{r}
# Pairs plot of the first few quant. variables
pairs(auto[, c("mpg", "cylinders", "displacement", "horsepower", "weight", "acceleration", "year")])

# Correlation matrix and heatmap
correlation_matrix <- cor(auto[, sapply(auto, is.numeric)])
corrplot(correlation_matrix, method = "circle")
```
- The graphical analysis reveals strong negative relationships between horsepower, weight, and displacement with mpg, indicating that cars with higher horsepower, greater weight, and larger engine displacement are significantly less fuel-efficient.
- In contrast, acceleration shows a weaker positive relationship with mpg, suggesting that vehicles with higher acceleration tend to have slightly better fuel efficiency. 

### 9. f)
Yes, the plots suggest that several variables might be useful in predicting mpg. 
The scatterplots reveal strong negative relationships between horsepower, weight, and displacement with mpg, indicating that these variables are likely significant predictors of fuel efficiency.
Hence, we see that horsepower, weight, and displacement are likely to be important variables in predicting mpg