---
title: "Week5_Ch4"
author: "Sahana Ramakrishnan"
date: "2024-08-14"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ISLR2)
library(dplyr)
library(MASS) # for LDA, QDA
library(class) # for KNN
library(e1071) # for naive bayes
library(ggplot2)
library(carat)
```

## Exercise 4.8
### 13. e)
```{r}
# loading weekly data
data("Weekly")
# Splitting data into training (1990-2008) and test (2009-2010) data sets
train <- Weekly[Weekly$Year <= 2008, ]
test <- Weekly[Weekly$Year >= 2009, ]

#LDA model using training data with Lag2 as predictor
lda_model_train <- lda(Direction ~ Lag2, data = train)
#predicting classes for test data
predicted_classes_test_lda <- predict(lda_model_train, newdata = test)$class

#test data confusion matrix
conf_matrix_test_lda <- table(Predicted = predicted_classes_test_lda, Actual = test$Direction)
conf_matrix_test_lda

#accuracy of test data
accuracy_test_lda <- sum(diag(conf_matrix_test_lda)) / sum(conf_matrix_test_lda)
accuracy_test_lda
```
### 13. f)
```{r}
#QDA model
qda_model_train <- qda(Direction ~ Lag2, data = train)

#predicting classes
predicted_classes_test_qda <- predict(qda_model_train, newdata = test)$class

#confusion matrix
conf_matrix_test_qda <- table(Predicted = predicted_classes_test_qda, Actual = test$Direction)
conf_matrix_test_qda

#accuracy
accuracy_test_qda <- sum(diag(conf_matrix_test_qda)) / sum(conf_matrix_test_qda)
accuracy_test_qda

```
### 13. g)
```{r}
#KNN method: Lag2 predictors in matrix format
train_X <- as.matrix(train$Lag2)
test_X <- as.matrix(test$Lag2)
train_Y <- train$Direction

#KNN with K=1
predicted_classes_test_knn <- knn(train = train_X, test = test_X, cl = train_Y, k = 1)

#confusion matrix
conf_matrix_test_knn <- table(Predicted = predicted_classes_test_knn, Actual = test$Direction)
conf_matrix_test_knn

#accuracy
accuracy_test_knn <- sum(diag(conf_matrix_test_knn)) / sum(conf_matrix_test_knn)
accuracy_test_knn

```
- knn() function from the class package, setting K=1 means that for each test observation, the class label of the nearest training observation is used to make predictions

### 13. h)
```{r}
# Naive bayes model using training data
nb_model_train <- naiveBayes(Direction ~ Lag2, data = train)

#predict classes
predicted_classes_test_nb <- predict(nb_model_train, newdata = test)

#confusion matrix
conf_matrix_test_nb <- table(Predicted = predicted_classes_test_nb, Actual = test$Direction)
conf_matrix_test_nb

#accuracy
accuracy_test_nb <- sum(diag(conf_matrix_test_nb)) / sum(conf_matrix_test_nb)
accuracy_test_nb

```
### 13. i)

- Based on the accuracy from each model, 
Naive Bayes: 0.5865
KNN(K=1): 0.5
QDA: 0.5865
LDA: 0.625

LDA provides the highest accuracy. Logistic regression as the method tried in previous week, 13. d) also provides accuracy of 0.625 which also proves to be one of the best methods.

### 13. j)
```{r}
#variables
train$Lag2_squared <- train$Lag2^2
train$Lag1_Lag2 <- train$Lag1 * train$Lag2
test$Lag2_squared <- test$Lag2^2
test$Lag1_Lag2 <- test$Lag1 * test$Lag2

# logistic regression
log_model_extended <- glm(Direction ~ Lag2 + Lag1 + Lag2_squared +
                            Lag1_Lag2, data = train, family = binomial)
predicted_probs_log <- predict(log_model_extended, newdata = test, type = "response")
predicted_classes_log <- ifelse(predicted_probs_log > 0.5, "Up", "Down")
conf_matrix_log <- table(Predicted = predicted_classes_log, Actual = test$Direction)
accuracy_log <- sum(diag(conf_matrix_log)) / sum(conf_matrix_log)
sprintf("Log. regression accuracy: %f", accuracy_log)

#LDA
lda_model_extended <- lda(Direction ~ Lag2 + Lag1 + Lag2_squared + Lag1_Lag2, data = train)
predicted_classes_lda <- predict(lda_model_extended, newdata = test)$class
conf_matrix_lda <- table(Predicted = predicted_classes_lda, Actual = test$Direction)
accuracy_lda <- sum(diag(conf_matrix_lda)) / sum(conf_matrix_lda)
sprintf("LDA accuracy: %f", accuracy_lda)

#QDA:
qda_model_extended <- qda(Direction ~ Lag2 + Lag1 + Lag2_squared + Lag1_Lag2, data = train)
predicted_classes_qda <- predict(qda_model_extended, newdata = test)$class
conf_matrix_qda <- table(Predicted = predicted_classes_qda, Actual = test$Direction)
accuracy_qda <- sum(diag(conf_matrix_qda)) / sum(conf_matrix_qda)
sprintf("QDA accuracy: %f", accuracy_qda)

#KNN for different K values (5, 20, 26)
train_X <- as.matrix(train[, c("Lag2", "Lag1", "Lag2_squared", "Lag1_Lag2")])
test_X <- as.matrix(test[, c("Lag2", "Lag1", "Lag2_squared", "Lag1_Lag2")])

predicted_classes_knn_5 <- knn(train = train_X, test = test_X, cl = train$Direction, k = 5)
conf_matrix_knn_5 <- table(Predicted = predicted_classes_knn_5, Actual = test$Direction)
accuracy_knn_5 <- sum(diag(conf_matrix_knn_5)) / sum(conf_matrix_knn_5)
sprintf("KNN 5 accuracy: %f", accuracy_knn_5)

predicted_classes_knn_20 <- knn(train = train_X, test = test_X, cl = train$Direction, k = 20)
conf_matrix_knn_20 <- table(Predicted = predicted_classes_knn_20, Actual = test$Direction)
accuracy_knn_20 <- sum(diag(conf_matrix_knn_20)) / sum(conf_matrix_knn_20)
sprintf("KNN 20 accuracy: %f", accuracy_knn_20)

predicted_classes_knn_25 <- knn(train = train_X, test = test_X, cl = train$Direction, k = 25)
conf_matrix_knn_25 <- table(Predicted = predicted_classes_knn_25, Actual = test$Direction)
accuracy_knn_25 <- sum(diag(conf_matrix_knn_25)) / sum(conf_matrix_knn_25)
sprintf("KNN 25 accuracy: %f", accuracy_knn_25)

#Naive Bayes
nb_model_extended <- naiveBayes(Direction ~ Lag2 + Lag1 + Lag2_squared + Lag1_Lag2, data = train)
predicted_classes_nb <- predict(nb_model_extended, newdata = test)
conf_matrix_nb <- table(Predicted = predicted_classes_nb, Actual = test$Direction)
accuracy_nb <- sum(diag(conf_matrix_nb)) / sum(conf_matrix_nb)
sprintf("Naive Bayes accuracy: %f", accuracy_nb)

```
- KNN with K=5 and Naive Bayes both achieved the highest accuracy of 0.586538
- Logistic Regression and LDA performed similarly with an accuracy of 0.557692
- QDA performed the least comparatively, with an accuracy of 0.471154
- KNN with K=20 and K=26 did not perform as well as K=5, suggesting that a smaller K might be more suitable for the dataset

- The best performing methods here are KNN with K=5 and Naive Bayes.
- If more interpretability is desired out of the dataset, the we can still consider using LDA despite the slightly lower accuracy.

### 14. a)
```{r}
#loading auto dataset
data("Auto")

#median of mpg
median_mpg <- median(Auto$mpg)

#binary variable mpg01
Auto$mpg01 <- ifelse(Auto$mpg > median_mpg, 1, 0)
```

### 14. b)
```{r}
#scatterplots
# mpg01 vs. displacement
ggplot(Auto, aes(x = displacement, y = mpg01)) + 
  geom_point() +
  labs(title = "mpg01 vs. displacement", x = "displacement", y = "mpg01")

# mpg01 vs. horsepower
ggplot(Auto, aes(x = horsepower, y = mpg01)) + 
  geom_point() +
  labs(title = "mpg01 vs. horsepower", x = "horsepower", y = "mpg01")

# mpg01 vs. weight
ggplot(Auto, aes(x = weight, y = mpg01)) + 
  geom_point() +
  labs(title = "mpg01 vs. weight", x = "weight", y = "mpg01")

# mpg01 vs. acceleration
ggplot(Auto, aes(x = acceleration, y = mpg01)) + 
  geom_point() +
  labs(title = "mpg01 vs. acceleration", x = "acceleration", y = "mpg01")

#Boxplots
# mpg01 vs. cylinders
ggplot(Auto, aes(x = factor(cylinders), y = mpg01)) + 
  geom_boxplot() +
  labs(title = "mpg01 vs. cylinders", x = "cylinders", y = "mpg01")

# mpg01 vs. year
ggplot(Auto, aes(x = factor(year), y = mpg01)) + 
  geom_boxplot() +
  labs(title = "mpg01 vs. year", x = "year", y = "mpg01")

# mpg01 vs. origin
ggplot(Auto, aes(x = factor(origin), y = mpg01)) + 
  geom_boxplot() +
  labs(title = "mpg01 vs. origin", x = "origin", y = "mpg01")
```
- There is a clear separation in mpg01 values with higher mpg01 values corresponding to lower displacement values
- There is a noticeable trend where lower weight corresponds to higher mpg01 values
- The boxplot shows that cars with fewer cylinders (3, 4, and 5) tend to have higher mpg01
- displacement, horsepower, weight, cylinders, year, and origin are the features most strongly associated with mpg01 and are likely to be useful predictors

### 14. d)
```{r}
# random numbers to reproduce
set.seed(120)
# no. of rows for train. data: 70% 
train_data_rows <- sample(1:nrow(Auto), 0.7 * nrow(Auto))

# Splitting into train. and test
train_set <- Auto[train_data_rows, ]
test_set <- Auto[-train_data_rows, ]

dim(train_set)
dim(test_set)

# choosing variables that closely associate with mpg01
selected_vars <- mpg01 ~ displacement + horsepower + weight + cylinders

#LDA on train. data
lda_model <- lda(selected_vars, data = train_set)

#predicting on test dataset
lda_predictions <- predict(lda_model, newdata = test_set)$class

#confusion matrix
conf_matrix_lda <- table(Predicted = lda_predictions, Actual = test_set$mpg01)
print("Confusion matrix: ")
conf_matrix_lda

#test error calculation
test_error_lda <- 1 - sum(diag(conf_matrix_lda)) / sum(conf_matrix_lda)
sprintf("Test error of the model: %f", test_error_lda)


```
- test error of the model is 0.135593 which means that approximately 13.56% of the test data was misclassified by the LDA model.
- the model seems to be doing a reasonably good job of predicting whether a car gets high or low gas mileage based on the variables selected

### 14. e)
```{r}
# QDA for same variables selected 
qda_model <- qda(selected_vars, data = train_set)

#predictions
qda_predictions <- predict(qda_model, newdata = test_set)$class

#confusion matrix
conf_matrix_qda <- table(Predicted = qda_predictions, Actual = test_set$mpg01)
print("Confusion matrix: ")
conf_matrix_qda

test_error_qda <- 1 - sum(diag(conf_matrix_qda)) / sum(conf_matrix_qda)
sprintf("Test error of the model: %f", test_error_qda)

```
- Both the LDA and QDA models have the same test error rate of approximately 13.56% with differences in the confusion matrix numbers.

### 14. g)
```{r}
#Naive Bayes 
nb_model <- naiveBayes(selected_vars, data = train_set)

#predictions
nb_predictions <- predict(nb_model, newdata = test_set)

#Confusion matrix
conf_matrix_nb <- table(Predicted = nb_predictions, Actual = test_set$mpg01)
print("Confusion Matrix:")
conf_matrix_nb

#test error
test_error_nb <- 1 - sum(diag(conf_matrix_nb)) / sum(conf_matrix_nb)
sprintf("Test error of the model: %f", test_error_nb)

```
All 3 models — LDA, QDA, and Naive Bayes, are performing similarly with a test error of approximately 13.56%. This consistency suggests that the feature selection is robust and that these models are equally suited to this particular classification task

### 14. h)
```{r}
selected_vars <- c("displacement", "horsepower", "weight", "cylinders")
#splitting train. and test data
train_indices <- sample(1:nrow(Auto), size = 0.7 * nrow(Auto))
train_set <- Auto[train_indices, ]
test_set <- Auto[-train_indices, ]
# data matrices for KNN
train_X <- as.matrix(train_set[, selected_vars])
test_X <- as.matrix(test_set[, selected_vars])
train_Y <- train_set$mpg01

# K values to test
ks <- 1:50  # K from 1 to 50

# KNN for each K using sapply
test_errors <- sapply(ks, function(k) {
    knn_predictions <- knn(train = train_X, test = test_X, cl = train_Y, k = k)
    mean(knn_predictions != test_set$mpg01)
})

# Plotting test errors vs. K
plot(ks, test_errors, type = "o", xlab = "K", ylab = "Test Error", main = "KNN Test Error vs. K")

# selecting the best K with lowest test error
best_k <- ks[which.min(test_errors)]
best_test_error <- min(test_errors)
print(paste("Best K is:", best_k, "with test error:", best_test_error))
```
- the best K for KNN model is 30, with a test error of approximately 14.4%. This is a strong result, indicating that the KNN model performs well with this value of K on the Auto dataset
